Dialect: A Compact Line Notation Based on SMILES

### 2022-03-25 manuscript.tex
Improves the flow of Goals, Design, and Tradeoffs.

Dialect's purpose is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. Serializations utility lies in enabling general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure.

Two concerns factor prominently in any serialization scheme: \textit{syntax} and \textit{semantics}. Syntax sets the vocabulary of symbols to be used and how they may be arranged. Semantics set the rules determining what various arrangements of symbols mean. Failure to fully specify syntax can lead to needless loss of data. Failure to fully specify semantics can result in needless data corruption. Semantics in particular play an important role in Dialect due the the inherently complex nature of molecular representations.

Beyond syntax and semantics, a serialization scheme should be useful. Utility in this context follows directly from the scope of supported molecular representations. Constrain this scope too much and the system will cover too few molecules to be interesting. Relax the constraints too much and the system could become verbose, difficult to implement, or both. Dialect aims squarely for the middle ground: a system of molecular representation that is specific enough to yield a compact language, but general enough to cover a large amount of interesting chemical space.

Dialect's molecular representation system is based on four mutually-exclusive components, each of which is constrained as follows:

\begin{itemize}
\item \textit{Constitution}. The atoms present in a molecule and how they are connected with each other through bonds. Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model (as described in detail later).
\item \textit{Conformation}. The kind of stereoisomerism resulting from restricted rotation about a bond. Dialect can encode those conformations arising from restricted rotation about double bonds, which includes both alkene and cumulene stereoisomerism.
\item \textit{Configuration}. The kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Dialect supports configurations for tetracoordinate atoms (those attached to four neighbors) having tetrahedral symmetry.
\item \textit{Delocalization}. A type of bonding relationship in which electrons are distributed over two or more bonds, among three or more atoms, or across some combination of atoms and bonds. Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
\end{itemize}

@@

Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes (which map to atoms) and a set of pairwise relationships between them called edges (which map to bonds). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

A Dialect molecule consists of a graph with zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is allowed. Dialect imposes no upper bound on the number of atoms or bonds. However, practical limitations related to memory, storage, and CPU time will likely limit the size of molecules in practice.

The edges of a Dialect molecular graph are defined as a set of unordered pairs of non-identical nodes. The practical consequences of this are threefold. First, \textit{loops} are not allowed. A loop is an edge joining a node to itself. Second, edges are \textit{undirected}. An undirected edge is equivalent to another edge whose terminals are swapped. An undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A. Third, \textit{parallel edges} are disallowed. Edges are said to be parallel if they connect the same two nodes.

A molecular graph contains one or more \textit{connected components}. A connected component is a graph in which any two vertices are connected to each other by at least one path. Dialect places no restriction on the number of connected components, although practical limitations are likely to impose one.

>>>>

Dialect's purpose is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This is useful because it enables general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile ("molfile")~\cite{ctfileFormats}.

Molecular serialization require careful attention to two complementary rule sets: \textit{syntax} and \textit{semantics}. Syntax defines the set of valid strings for a language, whereas semantics determine the meaning of valid strings. Incomplete specification of either syntax or semantics can lead to data loss and corruption. The semantics for Dialect are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

As a molecular serialization scheme, Dialect must balance two opposing factors: generality and complexity. If the chemical space Dialect spans is too broad, complete implementation could become impractical due to language complexity. But if the language is made too simple, the chemical space it spans may be of little interest. An important goal for Dialect is therefore to address the most interesting chemical space with the simplest possible syntax and semantics.

Toward this end, Dialect is built on four mutually-exclusive concepts:

\begin{itemize}
\item \textit{Constitution}, the atoms present in a molecule and how they are connected with each other through bonds. Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model.
\item \textit{Conformation}, the kind of stereoisomerism resulting from restricted rotation about a bond. Dialect support restricted rotation about double bonds leading to alkene isomerism.
\item \textit{Configuration}, the kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Dialect supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
\item \textit{Delocalization}, a type of bonding in which electrons are shared among three or more atoms, or across a combination of atoms and bonds. Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
\end{itemize}

@@

Dialect molecules inherit behavior from an underlying \textit{molecular graph}~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes and a set of pairwise relationships between them called edges. A molecular graph is a specialized graph onto which chemically-relevant metadata have been overlaid. The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively. Molecular graphs are ubiquitous in cheminformatics, appearing in both serialization formats in-memory data structures found in software toolkits.

A Dialect molecule is a molecular graph having zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is therefore allowed. Dialect imposes an upper bound on the number of nodes of $2^{32} - 1$ and the same limit for the number of edges. However, practical limitations related to memory, storage, and CPU time will likely further constrain these limits in practice.

The edges of a Dialect molecular graph are defined as a set of unordered pairs of non-identical nodes. The practical consequences of this are threefold. First, \textit{loops} are not allowed. A loop is an edge joining a node to itself. Second, edges are \textit{undirected}. An undirected edge is equivalent to another edge whose terminals are transposed. With one exception that will be discussed in detail, an undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A. Third, \textit{parallel edges} are disallowed. Edges are said to be parallel if they connect the same two nodes.

A molecular graph contains zero or more \textit{connected components}. A connected component is a graph in which any two vertices are connected to each other by at least one path. The maximum number of connected components is equal to the maximum node count, although practical factors may reduce this limit.
